Dynamic parameter estimation in the nested multinomial logit choice model

### 2021-08-11 paper/paper.tex
rev

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \affiliation for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
 \title{Dynamic parameter estimation in the nested multinomial logit choice model}
%% \tnotetext[label1]{}
 \author{Max Kapur\corref{cor1}}
 \ead{maxkapur@snu.ac.kr}
%% \ead[url]{home page}
\cortext[cor1]{Declarations of interest: none}
% \affiliation{organization={Seoul National University}, city={Seoul}, country={South Korea}}
 \address{Seoul National University, Department of Industrial Engineering, 1 Gwanak-ro, Gwanak-gu, Seoul 08826, Republic of Korea}
%% \fntext[label3]{}

\begin{abstract}
This paper considers the dynamic parameter estimation of the nested multinomial logit choice model under a tight computational constraint. I provide an approximation scheme for the model parameters that updates the current estimate according to an intuitive step rule. Stochastic convergence is guaranteed under the Robbins--Monro condition. Numerical simulations demonstrate the technique's feasibility.
\end{abstract}

@@

The present paper answers in the affirmative by providing a dynamic approximation scheme for the multinomial logit choice model and its nested variant in a tightly constrained computational environment called the data stream. When a new observation of a consumer choice arrives, the technique proposed here updates the current parameter estimate according to an intuitive rule with a sequence of decreasing step sizes. I show that stochastic convergence is guaranteed under the Robbins--Monro condition, and the convergence rate can be honed by tuning step parameters and utilizing a rolling average. I conclude by demonstrating the feasibility of the technique through numerical simulations.

@@

I propose the following \emph{MNL parameter update:}
\begin{align} \label{mnlparameterupdate}
\delta^{(t+1)} = \delta^{(t)} + \frac{\alpha}{t^r} Z^{(t)}
\end{align}
$\alpha >0$ and $r \in (0, 1]$ are step parameters. The intuition behind this parameter update is as follows. Seeing that $y_t$ was chosen from $\mathcal{A}_t$, we 
\begin{itemize}
\item Increase the estimated preferability of the item that was picked,
\item Decrease the estimated preferability of the items that were rejected, and
\item Leave unchanged the preferability of items not in the presented assortment.
\end{itemize}
Notice that $\sum_{i=1}^n \delta_i$ remains constant under the parameter update, because $\sum_i Z_i = 0$.

\begin{proposition}If the observations are drawn from an MNL choice model and each consideration set appears with nonzero probability, then the MNL parameter update converges in expectation to the true parameters $\delta$.
\end{proposition}

@@

\subsection{Why the parameter update converges}
Consider the multivariate root-finding problem \[\text{find } \theta^*: F(\theta^*) = 0\]
and the iterative scheme \[\theta^{(t+1)} = \theta^{(t)} + \frac{\alpha}{t^r} F\left(\theta^{(t)} \right)\]
where $\alpha >0$ and $r \in (0, 1]$ are step parameters. When $F: \mathbb{R}^n \mapsto \mathbb{R}^n$ is Lipschitz continuous and its Jacobian is negative definite, such an iterative scheme is known to converge to the root, if it exists (\citealp[][prop. 1.2.4]{bertsekas1999}; \citealp[][thm. 11.6]{nocedal2006}). A typical example occurs when using gradient ascent to maximize a strictly concave function whose gradient is $F$.

Moreover, if we replace $F(\theta)$ with an \emph{experiment} $\hat F(\theta)$ satisfying $\mathbb{E}\left[\hat F(\theta)\right] = F(\theta)$, then the same iterative scheme exhibits \emph{stochastic convergence} \cite[][]{robbinsmonro1951, tsitsiklis1994}. That is, with probability one,
\[\lim_{t\to \infty} \mathbb{E}\left[F\bigl(\theta^{(t)}\bigr)\right] = \theta^*\]
Many convergence rate results also generalize to stochastic interpretations \cite[][\S10.4]{kushner1997}.

@@

\subsection{The MNL parameter update is a Robbins--Monro algorithm}
To express the parameter update as a Robbins--Monro algorithm, it suffices to show that the optimal parameters are the zero of some function $F$, and that the parameter update accords with an experiment $\hat F$ which is alike in expectation.

Let $\delta$ denote the current parameter estimate and let $\delta^*$ denote the true, unknown parameters of the underlying MNL choice model. When the probability of offering each consideration set $\mathcal{A}$ is $\operatorname{Pr[\mathcal{A}]} > 0$, then the expected value of $Z$ has
\begin{align}
\mathbb{E}\left[Z(\delta)_i \right] &= \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\operatorname{Pr}\left[\mathcal{A}\right] \cdot \mathbb{E}\left[Z(\delta)_i~|~\mathcal{A}\right] \\
&=   \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\operatorname{Pr}\left[\mathcal{A}\right] \cdot \left(\operatorname{Pr}\left[\,\text{choose } i\;|\;\mathcal{A}\,\right] - \frac{\exp \delta_i}{\sum_{j\in \mathcal{A}} \exp \delta_j} \right) \\
&= \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\left(\frac{\exp  \delta_i^*}{\sum_{j\in \mathcal{A}} \exp \delta_j^*} - \frac{\exp \delta_i}{\sum_{j\in \mathcal{A}} \exp \delta_j} \right) \\
&= 0
\end{align}
if and only if $\delta^* = \delta$. Hence, the counterpart to $F(\theta)$ is the expected value of the random function $Z(\delta)$, while the counterpart to the experiment $\hat F(\theta)$ is the realization $Z^{(t)}(\delta)$. Next, observe that the Jacobian of $\mathbb{E}\left[Z(\delta) \right]$ is negative definite: The first term of the summand is constant, and the second is identical in form to the gradient of the logarithm of the likelihood function given in Equation \eqref{mnllikelihood}, which is concave. Thus, we conclude that the iterative procedure is a Robbins--Monro algorithm and converges in expectation to $\delta^*$.

\subsection{The NMNL parameter update}
The NMNL parameter update is analogous to the MNL parameter update. Each new observation consists of a consideration set $\mathcal{A}_t$, a choice of nest $x_t$, and a choice of product $y_t$. Define $\mathcal{A}_t^{\text{nests}}$ and $\mathcal{A}_{t}^{\text{products}}$ as above, and let
\begin{align}
W^{(t)}_k &= \begin{cases}
 -  \frac{\exp \sigma^{(t)}_k}{\sum_{l\in \mathcal{A}_{t}^{\text{nests}}} \exp \sigma_l}, \quad & k \in \mathcal{A}_{t}^{\text{nests}}, k \neq x_t\\
1 -  \frac{\exp \sigma^{(t)}_k}{\sum_{l\in \mathcal{A}_{t}^{\text{nests}}} \exp \sigma_l}, \quad & k \in \mathcal{A}_{t}^{\text{nests}}, k = x_t\\
0, \quad & k \notin \mathcal{A}_{t}^{\text{nests}}
\end{cases} \\
Z^{(t)}_{ki} &= 
\begin{cases}
 - \frac{\exp \delta^{(t)}_i}{\sum_{j\in \mathcal{A}_t^{\text{products}}} \exp \delta^{(t)}_j}, \quad&k = x_t, i \in \mathcal{A}_t^{\text{products}}, i\neq y_t \\
1 - \frac{\exp \delta^{(t)}_i}{\sum_{j\in \mathcal{A}_t^{\text{products}}} \exp \delta^{(t)}_j} , \quad&k = x_t, i \in \mathcal{A}_t^{\text{products}}, i =  y_t \\
0, \quad & \text{otherwise}
\end{cases}
\end{align}
The parameters are updated as follows.
\begin{align}\label{nmnlparameterupdate}
\sigma^{(t+1)} &= \sigma^{(t)} + \frac{\beta}{t^r} W^{(t)} \\
\delta^{(t+1)} &= \delta^{(t)} +  \frac{\alpha}{t^r} Z^{(t)} 
\end{align}
$\beta, \alpha >0$ and $r \in (0, 1]$ are step parameters.

>>>>

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \affiliation for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
 \title{Dynamic parameter estimation in the nested multinomial logit choice model}
%% \tnotetext[label1]{}
 \author{Max Kapur\corref{cor1}}
 \ead{maxkapur@snu.ac.kr}
%% \ead[url]{home page}
\cortext[cor1]{This research is supported by a grant from Seoul Metro, which manages the public transportation network in Seoul.}
% \affiliation{organization={Seoul National University}, city={Seoul}, country={South Korea}}
 \address{Seoul National University, Department of Industrial Engineering, 1 Gwanak-ro, Gwanak-gu, Seoul 08826, Republic of Korea}
%% \fntext[label3]{}

\begin{abstract}
This paper considers the dynamic parameter estimation of the nested multinomial logit choice model under a tight computational constraint. We provide an approximation scheme for the model parameters that updates the current estimate according to an intuitive step rule. Stochastic convergence is guaranteed under the Robbins--Monro condition. Numerical simulations demonstrate the technique's feasibility.
\end{abstract}

@@

The present paper answers in the affirmative by providing a dynamic approximation scheme for the multinomial logit choice model and its nested variant in a tightly constrained computational environment called the data stream. When a new observation of a consumer choice arrives, the technique proposed here updates the current parameter estimate according to an intuitive rule with a sequence of decreasing step sizes. We show that stochastic convergence is guaranteed under the Robbins--Monro condition. The convergence rate can be further improved by tuning step parameters and utilizing a rolling average. We conclude by demonstrating the feasibility of the technique through numerical simulations.

@@

We propose the following \emph{MNL parameter update:}
\begin{align} \label{mnlparameterupdate}
\delta^{(t+1)} = \delta^{(t)} + \frac{\alpha}{t^r} Z^{(t)}
\end{align}
$\alpha >0$ and $r \in (0, 1]$ are step parameters. The intuition behind this parameter update is as follows. Seeing that $y_t$ was chosen from $\mathcal{A}_t$, we 
\begin{itemize}
\item Increase the estimated preferability of the item that was picked,
\item Decrease the estimated preferability of items that were rejected, and
\item Leave unchanged the preferability of items not in the presented assortment.
\end{itemize}
Notice that $\sum_{i=1}^n \delta_i$ remains constant under the parameter update, because $\sum_i Z_i = 0$. The starting point $\delta^{(1)}$ may be chosen arbitrarily.

\begin{proposition}\label{mnlupdateconverges}If the observations are drawn from an MNL choice model and each consideration set appears with nonzero probability, then the MNL parameter update \eqref{mnlparameterupdate} converges in expectation to the true parameters $\delta$. 
\end{proposition}

@@

Before proving Proposition \ref{mnlupdateconverges}, we need to invoke a few canonical results.

\subsection{Convergence of gradient processes using decreasing step sizes}
Consider the multivariate root-finding problem \[\text{find } \theta^*: F(\theta^*) = 0\]
and the iterative scheme 
\begin{equation} \label{decreasingsteps}
\theta^{(t+1)} = \theta^{(t)} + \frac{\alpha}{t^r} F\left(\theta^{(t)} \right)
\end{equation}
where $\alpha >0$ and $r \in (0, 1]$ are step parameters. When $F: \mathbb{R}^n \mapsto \mathbb{R}^n$ is Lipschitz continuous and its Jacobian is negative definite, such an iterative scheme converges to the root, if it exists. A familiar example occurs when using gradient ascent to maximize a strictly concave function whose gradient is $F$ \cite[e.g.][prop. 1.2.4]{bertsekas1999}. The extension to root finding is addressed in theorem 11.6 of \cite{nocedal2006}, which shows that $F$ is a descent direction for the sum of squares $f(\theta) = \sum_{i=1}^n F_i(\theta)^2$, whose minimum occurs at the root.

\subsection{Stochastic convergence and Robbins--Monro algorithms}
Suppose that in the update step \eqref{decreasingsteps},  $F(\theta)$ is replaced by an \emph{experiment} $\hat F(\theta)$ satisfying $\mathbb{E}\left[\hat F(\theta)\right] = F(\theta)$. Then the same iterative scheme is called a \emph{Robbins--Monro algorithm}, and exhibits \emph{stochastic convergence}; that is, with probability one,\[\lim_{t\to \infty} \mathbb{E}\left[F\bigl(\theta^{(t)}\bigr)\right] = \theta^*\]
\cite{robbinsmonro1951} presented this result for univariate $F$; a more general result appears in \cite{tsitsiklis1994}. Many convergence rate results from the deterministic case also generalize to stochastic interpretations \cite[][\S10.4]{kushner1997}.

@@

\subsection{Proof of Proposition \ref{mnlupdateconverges}}
To prove that the MNL parameter update is a Robbins--Monro algorithm, it suffices to show that the optimal parameters are the zero of some function $F$ whose Jacobian is negative definite, and that the parameter update accords with an experiment $\hat F$ which is alike in expectation.

Let $\delta$ denote the current parameter estimate and let $\delta^*$ denote the true, unknown parameters of the underlying MNL choice model. When the probability of offering each consideration set $\mathcal{A}$ is $\operatorname{Pr[\mathcal{A}]} > 0$, then the expected value of $Z$ has, at each index,
\begin{align}
\mathbb{E}\left[Z(\delta)_i \right] &= \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\operatorname{Pr}\left[\mathcal{A}\right] \cdot \mathbb{E}\left[Z(\delta)_i~|~\mathcal{A}\right] \\
&=   \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\operatorname{Pr}\left[\mathcal{A}\right] \cdot \left(\operatorname{Pr}\left[\,\text{choose } i\;|\;\mathcal{A}\,\right] - \frac{\exp \delta_i}{\sum_{j\in \mathcal{A}} \exp \delta_j} \right) \\
&= \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\operatorname{Pr}\left[\mathcal{A}\right] \cdot \left(\frac{\exp  \delta_i^*}{\sum_{j\in \mathcal{A}} \exp \delta_j^*} - \frac{\exp \delta_i}{\sum_{j\in \mathcal{A}} \exp \delta_j} \right) \\
&= 0
\end{align}
if and only if $\delta^* = \delta$. Hence, the counterpart to $F(\theta)$ is the expected value of the random function $Z(\delta)$, while the counterpart to the experiment $\hat F(\theta)$ is the realization $Z^{(t)}(\delta)$. Next, observe that the Jacobian of $\mathbb{E}\left[Z(\delta) \right]$ is negative definite: The first term of the summand is constant, and the second is identical in form to the gradient of the logarithm of the likelihood function given in Equation \eqref{mnllikelihood}, which is concave. Thus, we conclude that the iterative procedure is a Robbins--Monro algorithm and converges in expectation to $\delta^*$.

\subsection{The NMNL parameter update}
The NMNL parameter update is analogous to the MNL parameter update. Each new observation consists of a consideration set $\mathcal{A}_t$, a nest choice $x_t$, and a product choice $y_t$. Define $\mathcal{A}_t^{\text{nests}}$ and $\mathcal{A}_{t}^{\text{products}}$ as above, and let
\begin{align}
W^{(t)}_k &= \begin{cases}
 -  \frac{\exp \sigma^{(t)}_k}{\sum_{l\in \mathcal{A}_{t}^{\text{nests}}} \exp \sigma_l}, \quad & k \in \mathcal{A}_{t}^{\text{nests}}, k \neq x_t\\
1 -  \frac{\exp \sigma^{(t)}_k}{\sum_{l\in \mathcal{A}_{t}^{\text{nests}}} \exp \sigma_l}, \quad & k \in \mathcal{A}_{t}^{\text{nests}}, k = x_t\\
0, \quad & k \notin \mathcal{A}_{t}^{\text{nests}}
\end{cases} \\
Z^{(t)}_{ki} &= 
\begin{cases}
 - \frac{\exp \delta^{(t)}_i}{\sum_{j\in \mathcal{A}_t^{\text{products}}} \exp \delta^{(t)}_j}, \quad&k = x_t, i \in \mathcal{A}_t^{\text{products}}, i\neq y_t \\
1 - \frac{\exp \delta^{(t)}_i}{\sum_{j\in \mathcal{A}_t^{\text{products}}} \exp \delta^{(t)}_j} , \quad&k = x_t, i \in \mathcal{A}_t^{\text{products}}, i =  y_t \\
0, \quad & \text{otherwise}
\end{cases}
\end{align}
The parameters are updated as follows.
\begin{align}\label{nmnlparameterupdate}
\sigma^{(t+1)} &= \sigma^{(t)} + \frac{\beta}{t^r} W^{(t)} \\
\delta^{(t+1)} &= \delta^{(t)} +  \frac{\alpha}{t^r} Z^{(t)} 
\end{align}
$\beta, \alpha >0$ and $r \in (0, 1]$ are step parameters.
